Limited plasticity but increased variance in physiological rates across ectotherm populations under climate change

### 2025-01-06 ms.qmd
realise I did not properly deal with R1 comment on aoverall so did that now

            contrast_lnrr_trait_cat_stats <- rbind(get_whole_bio(pred_lnrr_trait_post, habitat = "t", trait_b_data),
                                                        get_whole_bio(pred_lnrr_trait_post, habitat = "m", trait_b_data),
                                                        get_whole_bio(pred_lnrr_trait_post, habitat = "f", trait_b_data))
###########################################################
## Model 1.4 - whether chance between q10 depends on life stage and habitat
###########################################################
rerun = FALSE
if(rerun){
  # More realistic model with 
  model1.4 <- metafor::rma.mv(effect_size ~  max_temp_c + c_accl_time + type*habitat*life.hist, V = V, 
                              random = list(~1 + type | record_num,
                                            ~1 | species_full,
                                            ~1 | trait,
                                            ~1 + max_temp_c| spp,
                                            ~1 | obs),
                              R = list(species_full = phylo_cor), dfs = "contain", 
                              test = "t", data = lnRRQ10_data)

@@

@@

Acclimation capacity did not vary consistently by life-history stage with no differences in $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ between adult and juveniles, except for marine habitats where adults acclimated to a greater extent ("Model 4"; [@fig-cvrlh]A-C). Averaging over acute and acclimation effects there were also no differences between adults and juveniles within habitats (Adult-Juvenile: Terrestrial: `r contrast_lnrr_lh_cat_stats[1,3]`, 95\% CI: `r contrast_lnrr_lh_cat_stats[1,4]` to `r contrast_lnrr_lh_cat_stats[1,5]`, $p_{MCMC}$ = `r contrast_lnrr_lh_cat_stats[1,6]`; Marine: `r contrast_lnrr_lh_cat_stats[2,3]`, 95\% CI: `r contrast_lnrr_lh_cat_stats[2,4]` to `r contrast_lnrr_lh_cat_stats[2,5]`, $p_{MCMC}$ = `r contrast_lnrr_lh_cat_stats[2,6]`; Freshwater: `r contrast_lnrr_lh_cat_stats[3,3]`, 95\% CI: `r contrast_lnrr_lh_cat_stats[3,4]` to `r contrast_lnrr_lh_cat_stats[3,5]`, $p_{MCMC}$ = `r contrast_lnrr_lh_cat_stats[3,6]`).

@@

@@

@@

Each life-history stage exhibited the same pattern of variance change in each of the habitats (Adult-Juvenile contrasts: Marine: $\beta$ = `r contrast_lnvr_lh_cat_stats[2,3]`, 95\% CI: `r contrast_lnvr_lh_cat_stats[2,4]` to `r contrast_lnvr_lh_cat_stats[2,5]`, $p_{MCMC}$ = `r contrast_lnvr_lh_cat_stats[2,6]`; Freshwater: $\beta$ = `r contrast_lnvr_lh_cat_stats[3,3]`, 95\% CI: `r contrast_lnvr_lh_cat_stats[3,4]` to `r contrast_lnvr_lh_cat_stats[3,5]`, $p_{MCMC}$ = `r contrast_lnvr_lh_cat_stats[3,6]`; Terrestrial: $\beta$ = `r contrast_lnvr_lh_cat_stats[1,3]`, 95\% CI: `r contrast_lnvr_lh_cat_stats[1,4]` to `r contrast_lnvr_lh_cat_stats[1,5]`, $p_{MCMC}$ = `r contrast_lnvr_lh_cat_stats[1,6]`), with no differences between acute and acclimation effect types ("Model 4"; @fig-cvrlh).

@@

```{r, fig-delta}
#| label: fig-delta
#| fig-cap: "Meta-analysis results for $\\Delta lnRR_{Q_{10}}$ across different habitats. Thick bars are 95% confidence intervals (CI) and thin bars 95% prediction intervals (PI). $\\beta$ values are the contrasts between acute and acclimation means within each habitat. $\\mu$ values are the overall meta-analytic means averaged across acute and acclimation types within each habitat type. 95% CI's are indicated within square brackets and raw effects are weighted by their precision (inverse sampling variance). $p_{MCMC}$ values are the posterior probability of the contrast or overall meta-analytic mean being different from zero. k = total number of effect size estimates while the numbers in brackets indicate the number of species.  For ease of visualisation, all the raw data plotted for both acute and acclimation type effect sizes are presented as circles. Raw effects are weighted by their precision (inverse sampling variance). "

@@

Across habitats, the extent to which whole-organism versus biochemical traits acclimated varied ("Model 3"; [@fig-btrait]A-C). Biochemical traits acclimated to a greater extent compared to whole-organism traits in marine habitats ([@fig-btrait]B), whereas both whole-organism and biochemical traits acclimated similarly in freshwater ectotherms ([@fig-btrait]C). Neither trait category acclimated in terrestrial ectotherms ([@fig-btrait]A). However, there were no biochemical traits measured for juveniles in terrestrial species confounding life stage and trat category – though there were no differences between adult and juveniles in any case (see results in main manuscript).

Across habitats biochemical processes tended to result in greater increases in variance at higher temperatures, but this was only significant for marine ectotherms (biochemical/whole-organism contrasts: Marine: $\beta$ = `r contrast_lnvr_trait_cat_stats[2,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[2,4]` to `r contrast_lnvr_trait_cat_stats[2,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[2,6]`; Freshwater: $\beta$ = `r contrast_lnvr_trait_cat_stats[3,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[3,4]` to `r contrast_lnvr_trait_cat_stats[3,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[3,6]`; Terrestrial: $\beta$ = `r contrast_lnvr_trait_cat_stats[1,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[1,4]` to `r contrast_lnvr_trait_cat_stats[1,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[1,6]`) ([@fig-btrait]D-F; "Model 3"). Variance increases for biochemical traits was reduced during acclimation in marine ectotherms ([@fig-btrait]E).

>>>>

            contrast_lnrr_trait_cat_stats <- rbind(get_whole_bio(pred_lnrr_trait_post, habitat = "t", trait_b_data),
                                                        get_whole_bio(pred_lnrr_trait_post, habitat = "m", trait_b_data),
                                                        get_whole_bio(pred_lnrr_trait_post, habitat = "f", trait_b_data))

#######------------------------------------------------------------#########
# Add overall trait R1
#######------------------------------------------------------------#########
  # Tissue/whole-organism acute and acclimation RR
    pred_lnrr_post_overall_acute_tissue = c(pred_lnrr_trait_post[,2], pred_lnrr_trait_post[,6], pred_lnrr_trait_post[,10])
    pred_lnrr_post_overall_aclim_tissue = c(pred_lnrr_trait_post[,1], pred_lnrr_trait_post[,5], pred_lnrr_trait_post[,9])
     contrast_lnrr_tissue <- pred_lnrr_post_overall_aclim_tissue - pred_lnrr_post_overall_acute_tissue

      # Biochemical acute and acclimation RR
    pred_lnrr_post_overall_acute_biochem = c(pred_lnrr_trait_post[,4], pred_lnrr_trait_post[,8], pred_lnrr_trait_post[,12])
    pred_lnrr_post_overall_aclim_biochem = c(pred_lnrr_trait_post[,3], pred_lnrr_trait_post[,7], pred_lnrr_trait_post[,11])
        contrast_lnrr_biochem <- pred_lnrr_post_overall_aclim_biochem - pred_lnrr_post_overall_acute_biochem

#######------------------------------------------------------------#########

###########################################################
## Model 1.4 - whether chance between q10 depends on life stage and habitat
###########################################################
rerun = FALSE
if(rerun){
  # More realistic model with 
  model1.4 <- metafor::rma.mv(effect_size ~  max_temp_c + c_accl_time + type*habitat*life.hist, V = V, 
                              random = list(~1 + type | record_num,
                                            ~1 | species_full,
                                            ~1 | trait,
                                            ~1 + max_temp_c| spp,
                                            ~1 | obs),
                              R = list(species_full = phylo_cor), dfs = "contain", 
                              test = "t", data = lnRRQ10_data)

@@

#######------------------------------------------------------------#########  
# Add overall trait R1 - pool habitats
#######------------------------------------------------------------#########

 # Adult acute and acclimation RR - Note here we use rows because it's a different matrix calculation
    pred_lnrr_post_overall_acute_adult <- c(pred[4,], pred[8,], pred[12,]);
    pred_lnrr_post_overall_aclim_adult <- c(pred[3,], pred[7,], pred[11,])
                contrast_lnrr_adult_lh <- pred_lnrr_post_overall_aclim_adult - pred_lnrr_post_overall_acute_adult

  # Juvenile acute and acclimation RR - Note here we use rows because it's a different matrix calculation
    pred_lnrr_post_overall_acute_jeu <-  c(pred[2,], pred[6,], pred[10,])
    pred_lnrr_post_overall_aclim_jeu <- c(pred[1,], pred[5,], pred[9,])
                contrast_lnrr_jeu_lh <- pred_lnrr_post_overall_aclim_jeu - pred_lnrr_post_overall_acute_jeu

@@

Acclimation capacity did not vary consistently by life-history stage with no differences in $lnRR_{Q_{10_{acclimation}}}$ and $lnRR_{Q_{10_{acute}}}$ between adult and juveniles (overall contrast: `mean(c(contrast_lnrr_adult_lh,contrast_lnrr_jeu_lh))` 95\% CI: `r quantile(c(contrast_lnrr_adult_lh,contrast_lnrr_jeu_lh), 0.025)` to `r quantile(c(contrast_lnrr_adult_lh,contrast_lnrr_jeu_lh), 0.975)`, $p_{MCMC}$ = `r pmcmc(c(contrast_lnrr_adult_lh,contrast_lnrr_jeu_lh))`), except for marine habitats where adults acclimated to a greater extent ("Model 4"; [@fig-cvrlh]A-C). Averaging over acute and acclimation effects there were also no differences between adults and juveniles within habitats (Adult-Juvenile: Terrestrial: `r contrast_lnrr_lh_cat_stats[1,3]`, 95\% CI: `r contrast_lnrr_lh_cat_stats[1,4]` to `r contrast_lnrr_lh_cat_stats[1,5]`, $p_{MCMC}$ = `r contrast_lnrr_lh_cat_stats[1,6]`; Marine: `r contrast_lnrr_lh_cat_stats[2,3]`, 95\% CI: `r contrast_lnrr_lh_cat_stats[2,4]` to `r contrast_lnrr_lh_cat_stats[2,5]`, $p_{MCMC}$ = `r contrast_lnrr_lh_cat_stats[2,6]`; Freshwater: `r contrast_lnrr_lh_cat_stats[3,3]`, 95\% CI: `r contrast_lnrr_lh_cat_stats[3,4]` to `r contrast_lnrr_lh_cat_stats[3,5]`, $p_{MCMC}$ = `r contrast_lnrr_lh_cat_stats[3,6]`).

@@

#######------------------------------------------------------------#########
         # Add overall trait R1
#######------------------------------------------------------------#########
      # Tissue/whole-organism acute and acclimation VR
    pred_lnvr_post_overall_acute_tissue = c(pred_lnvr_traitb_post[,2], pred_lnvr_traitb_post[,6], pred_lnvr_traitb_post[,10])
    pred_lnvr_post_overall_aclim_tissue = c(pred_lnvr_traitb_post[,1], pred_lnvr_traitb_post[,5], pred_lnvr_traitb_post[,9])
      pooled_tissue = c(pred_lnvr_post_overall_acute_tissue, pred_lnvr_post_overall_aclim_tissue)

      # Biochemical acute and acclimation VR
    pred_lnvr_post_overall_acute_biochem = c(pred_lnvr_traitb_post[,4], pred_lnvr_traitb_post[,8], pred_lnvr_traitb_post[,12])
    pred_lnvr_post_overall_aclim_biochem = c(pred_lnvr_traitb_post[,3], pred_lnvr_traitb_post[,7], pred_lnvr_traitb_post[,11])
      pooled_biochem = c(pred_lnvr_post_overall_acute_biochem, pred_lnvr_post_overall_aclim_biochem)
#######------------------------------------------------------------#########

@@

#######------------------------------------------------------------#########
# Add overall trait R1
#######------------------------------------------------------------#########
  # Adult acute and acclimation VR
    pred_lnvr_post_overall_acute_adult = c(pred_lh_vr[4,], pred_lh_vr[8,], pred_lh_vr[12,])
    pred_lnvr_post_overall_aclim_adult = c(pred_lh_vr[3,], pred_lh_vr[7,], pred_lh_vr[11,])
      pooled_adult = c(pred_lnvr_post_overall_aclim_adult, pred_lnvr_post_overall_acute_adult)

  # Juvenile acute and acclimation VR
    pred_lnvr_post_overall_acute_jeu = c(pred_lh_vr[2,], pred_lh_vr[6,], pred_lh_vr[10,]); 
    pred_lnvr_post_overall_aclim_jeu = c(pred_lh_vr[1,], pred_lh_vr[5,], pred_lh_vr[9,]); 
      pooled_juev = c(pred_lnvr_post_overall_acute_jeu, pred_lnvr_post_overall_aclim_jeu)

@@

Each life-history stage exhibited the same pattern of variance change in each of the habitats (Adult-Juvenile contrasts: Marine: $\beta$ = `r contrast_lnvr_lh_cat_stats[2,3]`, 95\% CI: `r contrast_lnvr_lh_cat_stats[2,4]` to `r contrast_lnvr_lh_cat_stats[2,5]`, $p_{MCMC}$ = `r contrast_lnvr_lh_cat_stats[2,6]`; Freshwater: $\beta$ = `r contrast_lnvr_lh_cat_stats[3,3]`, 95\% CI: `r contrast_lnvr_lh_cat_stats[3,4]` to `r contrast_lnvr_lh_cat_stats[3,5]`, $p_{MCMC}$ = `r contrast_lnvr_lh_cat_stats[3,6]`; Terrestrial: $\beta$ = `r contrast_lnvr_lh_cat_stats[1,3]`, 95\% CI: `r contrast_lnvr_lh_cat_stats[1,4]` to `r contrast_lnvr_lh_cat_stats[1,5]`, $p_{MCMC}$ = `r contrast_lnvr_lh_cat_stats[1,6]`, overall across habitats: $\beta$ = `r mean(pooled_adult - pooled_juev)`, 95\% CI: `r quantile(pooled_adult - pooled_juev, 0.025)` to `r quantile(pooled_adult - pooled_juev, 0.975)`, $p_{MCMC}$ = `r pmcmc(pooled_adult - pooled_juev)`), with no differences between acute and acclimation effect types ("Model 4"; @fig-cvrlh).

@@

```{r, fig-delta}
#| label: fig-delta
#| fig-cap: "Meta-analysis results for $\\Delta lnRR_{Q_{10}}$ across different habitats. Thick bars are 95% confidence intervals (CI) and thin bars 95% prediction intervals (PI). $\\mu$ values are the overall meta-analytic means averaged across acute and acclimation types within each habitat type. 95% CI's are indicated within square brackets and raw effects are weighted by their precision (inverse sampling variance). $p_{MCMC}$ values are the posterior probability of the contrast or overall meta-analytic mean being different from zero. k = total number of effect size estimates while the numbers in brackets indicate the number of species.  For ease of visualisation, all the raw data plotted for both acute and acclimation type effect sizes are presented as circles. Raw effects are weighted by their precision (inverse sampling variance). "

@@

Across habitats, the extent to which whole-organism versus biochemical traits acclimated varied ("Model 3"; [@fig-btrait]A-C). Overall, there was no difference between the capacity for biochemical and tissue/whole-organism traits to plastically adjust (overall difference between acclimation and acute $lnRR_{Q_{10}}$ across habitats: `r mean(c(contrast_lnrr_tissue, contrast_lnrr_biochem))`, 95% CI: `r quantile(c(contrast_lnrr_tissue, contrast_lnrr_biochem), 0.025)` to `r quantile(c(contrast_lnrr_tissue, contrast_lnrr_biochem), 0.975)`, $p_{MCMC}$ = `r pmcmc(c(contrast_lnrr_tissue, contrast_lnrr_biochem))`). Biochemical traits acclimated to a greater extent compared to whole-organism traits in marine habitats ([@fig-btrait]B), whereas both whole-organism and biochemical traits acclimated similarly in freshwater ectotherms ([@fig-btrait]C). Neither trait category acclimated in terrestrial ectotherms ([@fig-btrait]A). However, there were no biochemical traits measured for juveniles in terrestrial species confounding life stage and trat category – though there were no differences between adult and juveniles in any case (see results in main manuscript).

Across habitats biochemical processes tended to result in greater increases in variance at higher temperatures, but not significantly so (overall contrast: `r mean(pooled_biochem - pooled_tissue)`, 95% CI:`r quantile(pooled_biochem - pooled_tissue, 0.025)` to `r quantile(pooled_biochem - pooled_tissue, 0.975)`,  $p_{MCMC}$ = `r pmcmc(pooled_biochem - pooled_tissue)`). However, within habitats there was a significant for marine ectotherms (biochemical/whole-organism contrasts: Marine: $\beta$ = `r contrast_lnvr_trait_cat_stats[2,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[2,4]` to `r contrast_lnvr_trait_cat_stats[2,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[2,6]`; Freshwater: $\beta$ = `r contrast_lnvr_trait_cat_stats[3,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[3,4]` to `r contrast_lnvr_trait_cat_stats[3,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[3,6]`; Terrestrial: $\beta$ = `r contrast_lnvr_trait_cat_stats[1,3]`, 95% CI: `r contrast_lnvr_trait_cat_stats[1,4]` to `r contrast_lnvr_trait_cat_stats[1,5]`, $p_{MCMC}$ = `r contrast_lnvr_trait_cat_stats[1,6]`) ([@fig-btrait]D-F; "Model 3"). Variance increases for biochemical traits was reduced during acclimation in marine ectotherms ([@fig-btrait]E).
